---
title: "Agents and Traits Tutorial"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Agents and Traits Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(artificialLifeR)
```

## Purpose

This tutorial introduces `create_agents()`. Artificial-life simulations often begin with a population of agents. Agents are simplified individuals with states, traits, and sometimes locations.

In this tutorial, you will learn how to create agents, inspect their traits, visualize their positions, and summarize trait variation.

## Create a population

```{r create-basic}
agents <- create_agents(
  n_agents = 60,
  seed = 10
)

head(agents)
```

## Inspect the output

```{r inspect-agents}
str(agents)
```

The main columns include:

| Column | Meaning |
|---|---|
| `agent` | Agent identifier |
| `x`, `y` | Position in the world |
| `energy` | Available energy |
| `speed` | Movement capacity |
| `efficiency` | Ability to convert resources into energy |
| `reproduction_threshold` | Energy needed for reproduction |
| `age` | Agent age |
| `alive` | Survival status |

## Plot agent positions

```{r agent-position-plot}
plot_alife_sim(
  agents,
  x = "x",
  y = "y",
  type = "point"
)
```

## Summarize traits

```{r summarize-traits}
summary(agents[, c("energy", "speed", "efficiency", "reproduction_threshold")])
```

## Measure trait diversity

```{r measure-efficiency}
measure_life_like_complexity(
  agents,
  trait_col = "efficiency"
)
```

## Compare populations

Create two populations with different trait variability.

```{r compare-trait-sd}
low_variation <- create_agents(
  n_agents = 60,
  trait_sd = 0.03,
  seed = 10
)

high_variation <- create_agents(
  n_agents = 60,
  trait_sd = 0.20,
  seed = 10
)

rbind(
  low_variation = measure_life_like_complexity(low_variation, trait_col = "efficiency"),
  high_variation = measure_life_like_complexity(high_variation, trait_col = "efficiency")
)
```

## Interpretation

Higher trait variability gives the population more initial diversity. In artificial-life models, variation matters because selection and adaptation require differences among individuals.

However, variation alone is not evolution. Evolution-like dynamics also require inheritance, differential persistence, and change across generations.

## Suggested exercises

- Increase `n_agents` and inspect how the output changes.
- Change `energy_mean` and observe the distribution of energy.
- Change `trait_sd` and compare efficiency summaries.
- Ask which traits might matter under resource competition.

## Key takeaway

`create_agents()` provides the starting population for artificial-life simulations. The agents are simplified units, but they make it possible to explore trait variation, energy, and individual-level differences.

