---
title: "Getting Started with artificialLifeR"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Getting Started with artificialLifeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(artificialLifeR)
```

## Overview

`artificialLifeR` is an educational R package for exploring artificial-life ideas through simplified simulations. The package focuses on agents, resources, reproduction, mutation, selection, population dynamics, and life-like complexity.

The central idea is:

> Life-like patterns can arise from local interaction, variation, inheritance, and environmental constraint.

The simulations are toy models. They are designed for teaching and conceptual exploration, not for realistic biological prediction.

## What the package helps you explore

| Model family | Main question | Core function |
|---|---|---|
| Agents and traits | What are the artificial individuals? | `create_agents()` |
| Resource competition | How do agents interact with an environment? | `simulate_resource_competition()` |
| Reproduction | How can offspring inherit traits? | `simulate_reproduction()` |
| Mutation | How does variation enter a population? | `simulate_mutation()` |
| Selection | How do some agents persist more than others? | `simulate_selection()` |
| Population dynamics | How does population size change over time? | `simulate_population_dynamics()` |
| Measurement | How can outputs be summarized? | `measure_life_like_complexity()` |

## First example: create agents

```{r first-agents}
agents <- create_agents(
  n_agents = 50,
  seed = 1
)

head(agents)
```

## Interpretation

Each row represents one artificial agent. The agents have positions, energy, and simple traits such as speed and efficiency.

These agents are not organisms. They are simplified units for exploring artificial-life ideas.

## Resource competition

```{r first-competition}
competition <- simulate_resource_competition(
  n_agents = 50,
  steps = 40,
  resource_regen = 0.20,
  seed = 2
)

head(competition$summary)
```

## Plot average energy

```{r first-energy-plot}
plot_alife_sim(
  competition$summary,
  x = "step",
  y = "mean_energy",
  type = "line"
)
```

## Measure life-like complexity

```{r first-measure}
measure_life_like_complexity(
  competition$agents,
  trait_col = "energy",
  time_col = "step"
)
```

## Population dynamics

```{r first-population}
pop <- simulate_population_dynamics(
  initial_population = 40,
  steps = 60,
  carrying_capacity = 100,
  seed = 3
)

head(pop$summary)
```

```{r first-pop-plot}
plot_alife_sim(
  pop$summary,
  x = "step",
  y = "population",
  type = "line"
)
```

## Suggested learning path

1. Start with `create_agents()` to understand traits and populations.
2. Use `simulate_resource_competition()` to explore agents in an environment.
3. Use `simulate_reproduction()` and `simulate_mutation()` to explore inheritance and variation.
4. Use `simulate_selection()` to explore fitness-based survival.
5. Use `simulate_population_dynamics()` to study population-level change.
6. Use `measure_life_like_complexity()` and `plot_alife_sim()` to summarize and visualize outputs.
7. Read the Theory Guide to connect the code to artificial life, emergence, origin-of-life research, and responsible interpretation.

## Responsible interpretation

It is better to say:

> The model illustrates life-like dynamics in a simplified artificial system.

than:

> The model creates or fully explains life.

