---
title: "Measuring Life-like Complexity Tutorial"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Measuring Life-like Complexity Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(artificialLifeR)
```

## Purpose

This tutorial introduces `measure_life_like_complexity()`. Measuring life-like organization is difficult because life is not a single variable. Artificial-life models may involve energy, traits, reproduction, variation, selection, population change, and environmental interaction.

The function provides educational summaries, not a universal life score.

## Measure agent traits

```{r measure-agents}
agents <- create_agents(
  n_agents = 60,
  seed = 1
)

measure_life_like_complexity(
  agents,
  trait_col = "efficiency"
)
```

## Measure resource competition

```{r measure-competition}
competition <- simulate_resource_competition(
  n_agents = 50,
  steps = 40,
  seed = 2
)

measure_life_like_complexity(
  competition$agents,
  trait_col = "energy",
  time_col = "step"
)
```

## Measure population dynamics

```{r measure-population}
pop <- simulate_population_dynamics(
  initial_population = 40,
  steps = 60,
  seed = 3
)

measure_life_like_complexity(
  pop$agents,
  trait_col = "efficiency",
  time_col = "step"
)
```

## Compare two scenarios

```{r measure-scenarios}
low_resource <- simulate_population_dynamics(
  initial_population = 40,
  steps = 60,
  resource_level = 0.5,
  seed = 3
)

high_resource <- simulate_population_dynamics(
  initial_population = 40,
  steps = 60,
  resource_level = 1.5,
  seed = 3
)

rbind(
  low_resource = measure_life_like_complexity(low_resource$agents, trait_col = "efficiency", time_col = "step"),
  high_resource = measure_life_like_complexity(high_resource$agents, trait_col = "efficiency", time_col = "step")
)
```

## Interpretation

The metrics summarize selected features of a trait or variable. Entropy can summarize diversity. Standard deviation can summarize variation. Temporal metrics summarize change over time.

These summaries are useful, but they do not prove that a system is alive.

## Good uses

Use the function to:

- compare trait variation;
- summarize energy changes;
- compare population scenarios;
- support visual interpretation;
- ask better questions about artificial-life dynamics.

## Poor uses

Do not use the function to claim:

- that life has been fully measured;
- that one toy model is truly alive;
- that entropy alone captures life-like organization;
- that the model replaces biological theory.

## Suggested exercises

- Measure `energy`, `efficiency`, and `speed` in the same simulation.
- Compare low and high mutation rates.
- Compare low and high resource levels.
- Ask which metric best matches the visual output.

## Key takeaway

`measure_life_like_complexity()` summarizes selected features of artificial-life outputs. It is an interpretive aid, not a final definition of life or complexity.

