---
title: "Population Dynamics Tutorial"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Population Dynamics Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(artificialLifeR)
```

## Purpose

This tutorial introduces `simulate_population_dynamics()`. Population dynamics are important in artificial life because individual reproduction and death can produce population-level patterns over time.

In this tutorial, you will run a population simulation, plot population size, compare resource levels, and interpret output carefully.

## Basic simulation

```{r pop-basic}
pop <- simulate_population_dynamics(
  initial_population = 40,
  steps = 80,
  carrying_capacity = 120,
  resource_level = 1.0,
  seed = 8
)

head(pop$summary)
```

## Plot population over time

```{r pop-plot}
plot_alife_sim(
  pop$summary,
  x = "step",
  y = "population",
  type = "line"
)
```

## Plot mean energy

```{r pop-energy-plot}
plot_alife_sim(
  pop$summary,
  x = "step",
  y = "mean_energy",
  type = "line"
)
```

## Compare resource levels

```{r pop-resource-compare}
low_resource <- simulate_population_dynamics(
  initial_population = 40,
  steps = 80,
  carrying_capacity = 120,
  resource_level = 0.5,
  seed = 8
)

high_resource <- simulate_population_dynamics(
  initial_population = 40,
  steps = 80,
  carrying_capacity = 120,
  resource_level = 1.5,
  seed = 8
)

data.frame(
  scenario = c("low resource", "high resource"),
  final_population = c(tail(low_resource$summary$population, 1), tail(high_resource$summary$population, 1)),
  final_mean_energy = c(tail(low_resource$summary$mean_energy, 1), tail(high_resource$summary$mean_energy, 1))
)
```

## Measure trait change

```{r pop-measure}
measure_life_like_complexity(
  pop$agents,
  trait_col = "efficiency",
  time_col = "step"
)
```

## Compare carrying capacity

```{r pop-capacity-compare}
small_capacity <- simulate_population_dynamics(
  initial_population = 40,
  steps = 80,
  carrying_capacity = 60,
  resource_level = 1.0,
  seed = 8
)

large_capacity <- simulate_population_dynamics(
  initial_population = 40,
  steps = 80,
  carrying_capacity = 160,
  resource_level = 1.0,
  seed = 8
)

data.frame(
  scenario = c("small capacity", "large capacity"),
  final_population = c(tail(small_capacity$summary$population, 1), tail(large_capacity$summary$population, 1)),
  max_population = c(max(small_capacity$summary$population), max(large_capacity$summary$population))
)
```

## Interpretation

Population dynamics emerge from repeated individual-level processes: energy gain, energy cost, death, reproduction, inheritance, mutation, and selection.

The population curve is not imposed directly. It arises from the simulation rules.

## Suggested exercises

- Change `initial_population`.
- Change `carrying_capacity`.
- Increase or decrease `resource_level`.
- Change `mutation_rate`.
- Ask whether the population stabilizes, grows, or declines.

## Responsible interpretation

This is a simplified educational model. It does not replace ecological, evolutionary, or demographic modeling. It illustrates how population-level patterns can arise from individual-level rules.

