---
title: "Reproduction and Mutation Tutorial"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Reproduction and Mutation Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(artificialLifeR)
```

## Purpose

This tutorial introduces `simulate_reproduction()` and `simulate_mutation()`. Reproduction and mutation are central to artificial life because they allow traits to persist and vary across generations.

The goal is to show how offspring inherit traits and how mutation introduces variation.

## Create agents

```{r repro-create}
agents <- create_agents(
  n_agents = 30,
  energy_mean = 1.8,
  seed = 4
)

head(agents)
```

## Simulate reproduction

```{r repro-basic}
offspring_pop <- simulate_reproduction(
  agents,
  mutation_rate = 0.10,
  mutation_sd = 0.03,
  seed = 4
)

nrow(agents)
nrow(offspring_pop)
head(offspring_pop)
```

## Interpretation

Agents with enough energy can produce offspring. Offspring inherit traits from parents, with optional mutation. This is a simplified representation of inheritance, not a full biological reproduction model.

## Apply mutation to one trait

```{r mutation-basic}
mutated <- simulate_mutation(
  agents,
  trait = "efficiency",
  mutation_rate = 0.50,
  mutation_sd = 0.08,
  lower = 0.01,
  upper = 1,
  seed = 5
)

head(mutated)
```

## Compare trait summaries

```{r mutation-summary}
rbind(
  original = measure_life_like_complexity(agents, trait_col = "efficiency"),
  mutated = measure_life_like_complexity(mutated, trait_col = "efficiency")
)
```

## Compare low and high mutation rates

```{r mutation-rate-compare}
low_mutation <- simulate_mutation(
  agents,
  trait = "efficiency",
  mutation_rate = 0.05,
  mutation_sd = 0.08,
  lower = 0.01,
  upper = 1,
  seed = 5
)

high_mutation <- simulate_mutation(
  agents,
  trait = "efficiency",
  mutation_rate = 0.80,
  mutation_sd = 0.08,
  lower = 0.01,
  upper = 1,
  seed = 5
)

rbind(
  low_mutation = measure_life_like_complexity(low_mutation, trait_col = "efficiency"),
  high_mutation = measure_life_like_complexity(high_mutation, trait_col = "efficiency")
)
```

## Interpretation of mutation

Mutation introduces trait variation. Without variation, selection has little material to act on. With too much mutation, inherited structure may become unstable.

A careful interpretation is:

> Mutation changes trait distributions in the artificial population.

not:

> Mutation alone creates biological evolution.

## Suggested exercises

- Change `mutation_sd` and observe trait variability.
- Mutate `speed` instead of `efficiency`.
- Increase initial energy and observe reproduction.
- Ask how inheritance and variation work together.

## Key takeaway

`simulate_reproduction()` and `simulate_mutation()` provide simplified tools for exploring inheritance and variation. Together, they help illustrate how artificial populations can change across generations.
