---
title: "Resource Competition Tutorial"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Resource Competition Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(artificialLifeR)
```

## Purpose

This tutorial introduces `simulate_resource_competition()`. Resource competition is a core artificial-life idea because agents must interact with an environment to maintain energy.

In this tutorial, you will run a competition simulation, inspect agent and resource outputs, plot average energy, compare resource regeneration settings, and interpret results carefully.

## Basic simulation

```{r resource-basic}
competition <- simulate_resource_competition(
  n_agents = 50,
  steps = 50,
  n_resources = 30,
  resource_regen = 0.20,
  seed = 2
)

names(competition)
```

## Inspect outputs

```{r inspect-resource-output}
head(competition$agents)
head(competition$resources)
head(competition$summary)
```

The result is a list with:

| Object | Meaning |
|---|---|
| `agents` | Agent states over time |
| `resources` | Resource locations and amounts over time |
| `summary` | Population-level summary by time step |

## Plot average energy

```{r resource-energy-plot}
plot_alife_sim(
  competition$summary,
  x = "step",
  y = "mean_energy",
  type = "line"
)
```

## Plot number alive

```{r resource-alive-plot}
plot_alife_sim(
  competition$summary,
  x = "step",
  y = "n_alive",
  type = "line"
)
```

## Compare resource regeneration

```{r resource-regen-compare}
low_regen <- simulate_resource_competition(
  n_agents = 50,
  steps = 50,
  resource_regen = 0.05,
  seed = 2
)

high_regen <- simulate_resource_competition(
  n_agents = 50,
  steps = 50,
  resource_regen = 0.40,
  seed = 2
)

rbind(
  low_regen = measure_life_like_complexity(low_regen$agents, trait_col = "energy", time_col = "step"),
  high_regen = measure_life_like_complexity(high_regen$agents, trait_col = "energy", time_col = "step")
)
```

## Compare population summaries

```{r resource-summary-compare}
data.frame(
  scenario = c("low regeneration", "high regeneration"),
  final_alive = c(tail(low_regen$summary$n_alive, 1), tail(high_regen$summary$n_alive, 1)),
  final_mean_energy = c(tail(low_regen$summary$mean_energy, 1), tail(high_regen$summary$mean_energy, 1)),
  final_total_resource = c(tail(low_regen$summary$total_resource, 1), tail(high_regen$summary$total_resource, 1))
)
```

## Interpretation

Resource regeneration changes environmental constraint. If resources regenerate slowly, agents may lose energy or die. If resources regenerate quickly, more agents may maintain energy.

This illustrates an artificial-life principle:

> Agent behavior cannot be understood separately from the environment.

## Suggested exercises

- Increase `movement_cost` and observe survival.
- Increase `n_agents` while keeping resources fixed.
- Change `consumption_rate` and compare average energy.
- Ask whether the environment is resource-rich or resource-limited.

## Responsible interpretation

This model illustrates resource competition in a simplified world. It is not a full ecological model, but it helps users explore energy, constraint, and agent-environment interaction.

