---
title: "Selection Tutorial"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Selection Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(artificialLifeR)
```

## Purpose

This tutorial introduces `simulate_selection()`. Selection is a process in which some agents persist or reproduce more than others because of differences in traits or fitness.

In this package, selection is simplified. It helps learners explore how population composition can change when survival depends on energy, efficiency, or a fitness proxy.

## Create agents

```{r selection-agents}
agents <- create_agents(
  n_agents = 80,
  seed = 6
)

summary(agents$efficiency)
```

## Apply selection

```{r selection-basic}
selected <- simulate_selection(
  agents,
  survival_fraction = 0.40,
  stochastic = FALSE,
  seed = 6
)

nrow(agents)
nrow(selected)
head(selected)
```

## Compare before and after selection

```{r selection-compare}
rbind(
  before = measure_life_like_complexity(agents, trait_col = "efficiency"),
  after = measure_life_like_complexity(selected, trait_col = "efficiency")
)
```

## Plot energy before and after

```{r selection-energy-table}
comparison <- data.frame(
  group = c(rep("before", nrow(agents)), rep("after", nrow(selected))),
  energy = c(agents$energy, selected$energy),
  index = seq_len(nrow(agents) + nrow(selected))
)

head(comparison)
```

## Stochastic selection

Selection is not always deterministic. In stochastic selection, higher-fitness agents have a better chance of being selected, but chance still matters.

```{r stochastic-selection}
stochastic_selected <- simulate_selection(
  agents,
  survival_fraction = 0.40,
  stochastic = TRUE,
  seed = 7
)

rbind(
  deterministic = measure_life_like_complexity(selected, trait_col = "fitness"),
  stochastic = measure_life_like_complexity(stochastic_selected, trait_col = "fitness")
)
```

## Interpretation

Selection changes the composition of the population. In deterministic selection, the highest-fitness agents are kept. In stochastic selection, fitter agents are more likely to persist, but lower-fitness agents may sometimes remain.

This illustrates a key artificial-life idea:

> Population-level change can arise when individual differences affect survival or reproduction.

## Suggested exercises

- Change `survival_fraction` and compare results.
- Use deterministic and stochastic selection.
- Create a custom fitness column and select using `fitness_col`.
- Ask which traits become more common after selection.

## Responsible interpretation

This function illustrates selection in a simplified artificial system. It does not model full natural selection, genetics, development, or ecology.

